Using Fundamental Measure Theory to Treat the Correlation Function of the Inhomogeneous Hard-Sphere Fluid

### 2012-08-14 papers/contact/paper.tex
changed appendix equations in paper so that they sem to make more sense and are easier to match up with code Tested everything on node02.
fix minus-sign bug in A correlation, and use new (fixed) data (mark II correlation function still buggy) Tested everything on node06.
new title and abstract for contact paper (and new reference in water paper) Tests skipped!

\begin{document}
\title{A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities}

\author{Jeff Schulte}
\author{Chris Haglund}
\author{Patrick Kreitzburg}
\author{David Roundy}
\affiliation{Department of Physics, Oregon State University, Corvallis, OR 97331}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We investigate the contact density of an inhomogeneous hard-sphere
  fluid, which is of particular interest, as it plays a critical role
  in Statistical Associating Fluid Theory (SAFT), which is the basis
  of a number of recent classical density functionals.  We derive a
  formula for the contact density from the White Bear version of the
  Fundamental Measure Theory functional~\cite{roth2002whitebear}, and
  test this functional agains Monte Carlo simulations.
  \textcolor{red}{Insert summary of results here...}
\end{abstract}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/walls-10}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.1.}
  \label{fig:walls-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-4-10}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:inner-4-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-12-10}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:inner-12-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/outer-12-01}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:outer-10}
\end{figure}

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr) &= \int n(\rr') \Theta(\left|\rr - \rr'\right| -R(\rr')) d\rr'
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we needS
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta R(\rr)} &=
  \int n (\rr'') \delta(|\rr' - \rr''| - R(\rr'')) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - R(\rr))
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - R(\rr'')) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta R(\rr)} &= -n(\rr) \delta'(|\rr' - \rr| - R(\rr))\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{4\pi R(\rr'')}
  \delta(|\rr'-\rr''| - R(\rr'')) \\
  \frac{\delta n_1(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)}\delta'(|\rr'-\rr| - R(\rr))
  -
  \frac{n(\rr)}{4\pi
    R(\rr)^2}\delta(|\rr'-\rr| - R(\rr))
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{4\pi R(\rr'')^2}
  \delta(|\rr'-\rr''| - R(\rr'')) \\
  \frac{\delta n_0(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)^2}\delta'(|\rr'-\rr| - R(\rr))
  -
  2\frac{n(\rr)}{4\pi
    R(\rr)^3}\delta(|\rr'-\rr| - R(\rr))
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr) &= \int n(\rr') \delta(|\rr' - \rr''| - R(\rr''))
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta R(\rr)} &= - n(\rr) \delta'(|\rr' - \rr| - R(\rr))
    \frac{\rr'-\rr}{|\rr'-\rr|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr) &= \int d\rr' \frac{n(\rr')}{4\pi R(\rr')}
  \delta(|\mathbf{r}-\rr'| - R(\rr')) \frac{\rr-\rr'}{|\rr-\rr'|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)}\delta'(|\rr'-\rr| - R(\rr)) \frac{\rr'-\rr}{|\rr'-\rr|}
  -
  \frac{n(\rr)}{4\pi
    R(\rr)^2}\delta(|\rr'-\rr| - R(\rr)) \frac{\rr'-\rr}{|\rr'-\rr|}
\end{align}
\fixme{double-check the directions of all the vector quantities...}

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^S(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position at which the weighted
density is being evaluated.  This is less intuitive than the $A$ case,
since \emph{nothing is there}!  However, in practice it corresponds to
the picture shown in Figure~\ref{fig:contact}, which involves
extending the radius for spheres just at the position touching a
given spot, where we are evaluating the correlation function.  We
define the weighted density $n_3^S$ as:
\begin{align}
  n_3^{S}(\rr) &= \int n(\rr') \Theta(\left|\rr - \rr'\right| -R(\rr)) d\rr'
\end{align}
In the symmetric case, the functional derivative is strikingly simple:
\begin{align}
  \frac{\delta n_3^{S} (\rr')}{\delta R(\rr)} &=
   \delta(\rr'-\rr) \int n (\rr'') \delta(|\rr' - \rr''| - R(\rr'))
   \delta(\rr'-\rr) d\rr''
   \\
   &= \delta(\rr'-\rr) n_2^S(\rr')
\end{align}
The functional derivative does \emph{not} eliminate the integral,
which simply changes from one weighted density to another.  The
corresponding term in the free energy functional, once integrated,
takes a simple form:
\begin{align}
  \int \frac{\delta A_{HS}}{\delta n_3^{S}(\rr')}
  \frac{\delta n_3^{S}(\rr')}{\delta R(\rr)} d\rr'
  &= \int \frac{\delta A_{HS}}{\delta n_3^{S}(\rr')} \delta(\rr'-\rr)
  n_2^S(\rr') d\rr' \\
  &= \frac{\delta A_{HS}}{\delta n_3^{S}(\rr)} n_2(\rr)
\end{align}
which means that we can compute $g_\sigma^S(\rr)$ using only
convolutions of the density, in strong contrast with
$g_\sigma^A(\rr)$, which requires convolutions of derivatives of the
free energy density.

The functional derivative of $n_2^S(\rr')$ is quite similar:
\begin{align}
  n_2^{S}(\rr) &= \int n(\rr') \delta(|\rr - \rr'| - R(\rr))d\mathbf
  r'\\
  \frac{\delta n_2^{S}(\rr')}{\delta R(\rr)} &= -\delta(\rr-\rr') \int n(\rr'')
  \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
\end{align}
where we recognize that the convolution---while not one of the standard
FMT weighted density---is a simple convolution of the density.  We can
go on to compute the derivatives of the remaining weighted densities:
\begin{align}
  n_1^{S}(\rr) &= \int \frac{n(\rr')}{4\pi R(\rr)} \delta(|\rr - \rr'| - R(\rr))d\mathbf r'\\
    \frac{\delta n_1^{S}(\rr')}{\delta R(\rr)} &=
    -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi R(\rr)}
    \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
    + n_0^s(\rr') \right)
\end{align}
\begin{align}
  n_0^{S}(\rr) &= \int \frac{n(\rr')}{4\pi R(\rr)^2} \delta(|\rr - \rr'| - R(\rr))d\mathbf r'\\
    \frac{\delta n_0^{S}(\rr')}{\delta R(\rr)}
    &= -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi R(\rr)^2}
    \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
    +
    \frac{2}{R} n_0^S(\rr') \right)
\end{align}
\begin{align}
  \mathbf{n}_{V2}^{S}(\rr) &= \int n(\rr') \delta(|\rr - \rr'| - R(\rr))
  \frac{(\rr - \rr')}{|\rr - \rr'|} d \rr'\\
  \frac{\delta \mathbf{n}_{V2}^{S}(\rr')}{\delta R(\rr)} &= -\delta(\rr - \rr')
  \int n(\rr'') \delta'(|\rr' - \rr''| - R(\rr'))
  \frac{(\rr - \rr'')}{|\rr - \rr''|} d\rr''
\end{align}
\begin{align}
 \mathbf{n}_{V1}^{S}(\rr) &= \int \frac{n(\rr')}{4\pi R(\rr)} \delta(|\rr - \rr'| + R(\rr))d\mathbf r'
   \frac{(\rr - \rr')}{|\rr - \rr'|}\\
 \frac{\delta \mathbf{n}_{V1}^{S}(\rr')}{\delta R(\rr)}
 &= -\delta(\rr-\rr')\left(\int \frac{n(\rr'')}{4\pi R(\rr)}
   \delta'(|\rr'-\rr''| - R(\rr')) \frac{(\rr - \rr'')}{|\rr - \rr''|}  d\rr''
   + \frac{\mathbf{n}_{V2}^S(\rr')}{4\pi R(\rr)^2} \right)
\end{align}

>>>>

\begin{document}
\title{Using Fundamental Measure Theory to Treat the Correlation
  Function of the Inhomogeneous Hard-Sphere Fluid}

\author{Jeff Schulte}
\author{Patrick Kreitzburg}
\author{Chris Haglund}
\author{David Roundy}
\affiliation{Department of Physics, Oregon State University, Corvallis, OR 97331}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We investigate the value of the correlation function an
  inhomogeneous hard-sphere fluid at contact $g_\sigma$.  This
  quantity plays a critical role in Statistical Associating Fluid
  Theory (SAFT), which is the basis of a number of recent classical
  density functionals.  We define two averaged values for the
  correlation function at contact, and derive formulas for each of
  them White Bear version of the Fundamental Measure Theory
  functional~\cite{roth2002whitebear} based on an assumption of
  thermodynamic consistency. We have tested these functionals and two
  existing functionals~\cite{yu2002fmt-dft-inhomogeneous-associating,
    gross2009density} against Monte Carlo simulations.  We find that
  excellent agreement with the Monte Carlo results for our
  asymmetrically averaged correlation function, and less impressive
  results for the two existing functionals and our symmetrically
  averaged correlation function.
\end{abstract}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/walls-10}
  \caption{Density and contact densities at a wall with bulk filling
    fraction of 0.1.}
  \label{fig:walls-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-4-10}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:inner-4-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/inner-12-10}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:inner-12-10}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/outer-16-01}
  \caption{Density and contact densities at a xxx with bulk filling
    fraction of 0.1.}
  \label{fig:outer-10}
\end{figure}

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr') &= \int n(\rr'') \Theta(\left|\rr' - \rr''\right| -R(\rr'')) d\rr''
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we needS
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta R(\rr)} &=
  \int n (\rr'') \delta(|\rr' - \rr''| - R(\rr'')) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - R(\rr))
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - R(\rr'')) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta R(\rr)} &= -n(\rr) \delta'(|\rr' - \rr| - R(\rr))\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{4\pi R(\rr'')}
  \delta(|\rr'-\rr''| - R(\rr'')) \\
  \frac{\delta n_1(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)}\delta'(|\rr'-\rr| - R(\rr))
  -
  \frac{n(\rr)}{4\pi
    R(\rr)^2}\delta(|\rr'-\rr| - R(\rr))
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{4\pi R(\rr'')^2}
  \delta(|\rr'-\rr''| - R(\rr'')) \\
  \frac{\delta n_0(\rr')}{\delta R(\rr)}
  &= -\frac{n(\rr)}{4\pi
    R(\rr)^2}\delta'(|\rr'-\rr| - R(\rr))
  -
  2\frac{n(\rr)}{4\pi
    R(\rr)^3}\delta(|\rr'-\rr| - R(\rr))
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr') &= \int n(\rr') \delta(|\rr' - \rr''| - R(\rr''))
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta R(\rr)} &= n(\rr) \delta'(|\rr' - \rr| - R(\rr))
    \frac{\rr-\rr'}{|\rr-\rr'|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr') &= \int d\rr'' \frac{n(\rr'')}{4\pi R(\rr'')}
  \delta(|\mathbf{r'}-\rr''| - R(\rr'')) \frac{\rr'-\rr''}{|\rr'-\rr''|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta R(\rr)}
  &= \frac{n(\rr)}{4\pi
    R(\rr)}\delta'(|\rr'-\rr| - R(\rr)) \frac{\rr-\rr'}{|\rr-\rr'|}
  +
  \frac{n(\rr)}{4\pi
    R(\rr)^2}\delta(|\rr'-\rr| - R(\rr)) \frac{\rr-\rr'}{|\rr-\rr'|}
\end{align}
\fixme{double-check the directions of all the vector quantities...}

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^S(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position at which the weighted
density is being evaluated.  This is less intuitive than the $A$ case,
since \emph{nothing is there}!  However, in practice it corresponds to
the picture shown in Figure~\ref{fig:contact}, which involves
extending the radius for spheres just at the position touching a
given spot, where we are evaluating the correlation function.  We
define the weighted density $n_3^S$ as:
\begin{align}
  n_3^{S}(\rr') &= \int n(\rr'') \Theta(\left|\rr' - \rr''\right| -R(\rr')) d\rr''
\end{align}
In the symmetric case, the functional derivative is strikingly simple:
\begin{align}
  \frac{\delta n_3^{S} (\rr')}{\delta R(\rr)} &=
   \delta(\rr'-\rr) \int n (\rr'') \delta(|\rr' - \rr''| - R(\rr'))
   \delta(\rr'-\rr) d\rr''
   \\
   &= \delta(\rr'-\rr) n_2^S(\rr')
\end{align}
The functional derivative does \emph{not} eliminate the integral,
which simply changes from one weighted density to another.  The
corresponding term in the free energy functional, once integrated,
takes a simple form:
\begin{align}
  \int \frac{\delta A_{HS}}{\delta n_3^{S}(\rr')}
  \frac{\delta n_3^{S}(\rr')}{\delta R(\rr)} d\rr'
  &= \int \frac{\delta A_{HS}}{\delta n_3^{S}(\rr')} \delta(\rr'-\rr)
  n_2^S(\rr') d\rr' \\
  &= \frac{\delta A_{HS}}{\delta n_3^{S}(\rr)} n_2(\rr)
\end{align}
which means that we can compute $g_\sigma^S(\rr)$ using only
convolutions of the density, in strong contrast with
$g_\sigma^A(\rr)$, which requires convolutions of derivatives of the
free energy density.

The functional derivative of $n_2^S(\rr')$ is quite similar:
\begin{align}
  n_2^{S}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - R(\rr'))d\mathbf
  r''\\
  \frac{\delta n_2^{S}(\rr')}{\delta R(\rr)} &= -\delta(\rr-\rr') \int n(\rr'')
  \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
\end{align}
where we recognize that the convolution---while not one of the standard
FMT weighted density---is a simple convolution of the density.  We can
go on to compute the derivatives of the remaining weighted densities:
\begin{align}
  n_1^{S}(\rr') &= \int \frac{n(\rr'')}{4\pi R(\rr')} \delta(|\rr' - \rr''| - R(\rr'))d\mathbf r''\\
    \frac{\delta n_1^{S}(\rr')}{\delta R(\rr)} &=
    -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi R(\rr)}
    \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
    + n_0^s(\rr') \right)
\end{align}
\begin{align}
  n_0^{S}(\rr') &= \int \frac{n(\rr'')}{4\pi R(\rr')^2} \delta(|\rr' - \rr''| - R(\rr'))d\mathbf r''\\
    \frac{\delta n_0^{S}(\rr')}{\delta R(\rr)}
    &= -\delta(\rr-\rr') \left( \int \frac{n(\rr'')}{4\pi R(\rr)^2}
    \delta'(|\rr'-\rr''| - R(\rr')) d\rr''
    +
    \frac{2}{R} n_0^S(\rr') \right)
\end{align}
\begin{align}
  \mathbf{n}_{V2}^{S}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - R(\rr'))
  \frac{(\rr' - \rr'')}{|\rr' - \rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{S}(\rr')}{\delta R(\rr)} &= -\delta(\rr - \rr')
  \int n(\rr'') \delta'(|\rr' - \rr''| - R(\rr'))
  \frac{(\rr - \rr'')}{|\rr - \rr''|} d\rr''
\end{align}
\begin{align}
 \mathbf{n}_{V1}^{S}(\rr') &= \int \frac{n(\rr'')}{4\pi R(\rr')} \delta(|\rr' - \rr''| - R(\rr'))d\mathbf r''
   \frac{(\rr' - \rr'')}{|\rr' - \rr''|}\\
 \frac{\delta \mathbf{n}_{V1}^{S}(\rr')}{\delta R(\rr)}
 &= -\delta(\rr-\rr')\left(\int \frac{n(\rr'')}{4\pi R(\rr)}
   \delta'(|\rr'-\rr''| - R(\rr')) \frac{(\rr - \rr'')}{|\rr - \rr''|}  d\rr''
   + \frac{\mathbf{n}_{V2}^S(\rr')}{4\pi R(\rr)^2} \right)
\end{align}
